Testing for measurement invariance with respect to an ordinal variable

### 2012-08-30 ordinal/ordinal.Rnw
Finished full draft of ordinal manuscript: o Added figures to application section. o Added to general discussion. o Added references to bib file. o Added saved simulation results for fast compiling.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@167 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{Testing for measurement invariance with respect to an ordinal variable}
\threeauthors{Edgar C. Merkle}{Jinyan Fan}{Achim Zeileis}
\threeaffiliations{University of Missouri}{Auburn University}{Universit\"{a}t Innsbruck}

@@

As demonstrated in the simulations below, the ordinal test
statistics proposed above are sensitive to the measurement invariance
violations that an analyst would typically expect from an ordinal $V$.
In particular, due to computing cumulative sums in ${\bm B}(\hat {\bm \theta})$,
violations that occur as we move along the levels of $V$ can be captured
wll. This includes abrupt shifts in the parameters $\bm \theta$ at a
certain level of $V$ as well as smooth increases/decreases in the parameters.
Taking a maximum over the $k$~parameters as in $\mathit{WDM}_o$ will be
more sensitive to changes that occur only in one out of many
parameters, 
while $\max \mathit{LM}_o$ will be more sensitive to changes occuring
in several (or even all of the) parameters simultaneously.
Moreover, the test statistics are rather insensitive to anomolies in a small
number of categories of $V$ that are
unrelated to the ordering of $V$.  This is especially relevant to situations
where the analyst has a large sample size, where the usual
likelihood ratio test is notoriously sensitive to minor parameter
instabilities (e.g., \citeNP{Ben80}).

@@

In this simulation, we demonstrate that the proposed test statistics are
sensitive to ordinal measurement invariance violations, moreso than
traditional statistics.  We generate data from a two-factor,
six-indicator model, with a measurement invariance violation occurring
in the three unique variances associated with the first factor.  We
use the proposed test statistics to test for measurement invariance
simultaneously across all six unique variances, which mimics the
comparison of a tau-equivalent model to a congeneric model (e.g.,
\citeNP{VanLan00}).

For comparison, we also compute two statistics that treat the ordinal
auxiliary variable as categorical.  These include the likelihood ratio
test of measurement 
invariance in the six unique variance parameters, along with the
unordered LM test from~\eqref{eq:lmuo}.

@@

\section{Simulation 2: Minor Anomolies and Large $n$}

@@

\subsection{Results}
Simulation results are presented in Figure~\ref{fig:sim2res}.  It is
observed that results are very consistent across the sample sizes
tested, implying that ``practical infinity'' is reached for this model
by $n=1200$. \readme{Also implying that we don't need all three sample
  sizes or could try values smaller than 1200.}  We also observe
a negative 
relationship between power and $m$; this is because the measurement
invariance violation occurred at only one level of the auxiliary
variable.  As $m$ increases (and $n$ is held
constant), the number of individuals violating measurement invariance
therefore decreases.  As a result, power to detect the violation decreases with
increasing $m$.

@@

Taken together, the results from Simulation 1 and Simulation
2 provide evidence that the $\text{LM}_o$ statistic should be
preferred to the $\text{WDM}_o$ statistic for testing measurement
invariance. \readme{This is not surprising given that the models
have quite a few parameters and we change many of them. We could 
try to set up a simulation where only a single out of many parameters
changes where the WDM should perform better. Also,
in models that are simpler than most CFA models, the WDM may be more useful.
I'm not sure how much we want to comment on this or illustrate this.}
The $\text{LM}_o$ statistic has higher power to detect
ordinal violations, and its power to detect non-ordinal violations was
similar to that of $\text{WDM}_o$ when the violation magnitude was
small (e.g., for $d \leq 1.5$).  The $\text{LM}_o$ statistic is
advantageous because it can make use of invariance violations that
simultaneously 
occur in multiple parameters, whereas the $\text{WDM}_o$ focuses only
on the parameter with the largest invariance violation.  
Thus, the statistics are likely to exhibit
similar performance if only a single model parameter violated
measurement invariance.

@@

\begin{figure}
\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
  the $\text{LM}_o$ statistic (right panel).}
\label{fig:gq6res}
\setkeys{Gin}{width=\textwidth}
<<gq6res, fig=TRUE, height=3, width=6>>=
par(mfrow=c(1,2), cex=.65)

@@

Plots representing the statistics' fluctuations across levels of age
group are displayed in Figure~\ref{fig:gq6res}.  The left panel
displays the process associated with $\text{WDM}_o$, while the right
panel displays the process associated with $\text{LM}_o$.  The final
age group (17-19 years) is not displayed, because the statistics
associated with this final age group are always zero (scores are
summed across all individuals in the sample at this point).
%% more detail

@@

@@

\subsection{Summary}
The application considered above shows that the ordinal test
statistics can provide useful information in situations where one
might question significant likelihood ratio test statistics.  While
researchers use rules of thumb to obtain decisions from other
alternative fit measures, the proposed statistics are proper tests of
the hypothesis of interest.  They can be used to either supplement or
replace the likelihood ratio test, depending upon the measurement
invariance violations in which the researcher is interested a priori.
We further describe the issue of supplementing vs.\ replacing the
likelihood ratio test in the general discussion.

@@

@@

\subsection{Summary}
The test statistics proposed in this paper have relatively-high power for
detecting measurement 
invariance violations that are monotonic with the ordinal variable,
and they have relatively-low power for detecting minor violations that are not
monotonic.  The former feature implies that the statistics are good at
detecting measurement invariance violations that are interpretable to
the researcher, while the latter feature implies that the statistics
are feasible in situations where the likelihood ratio test commonly
rejects $H_0$ in practice (e.g., \citeNP{Ben80}).  Furthermore, the
focal psychometric model does not have to be modified in any way,
which differs from approaches that may treat the ordinal variable as
continuous 
(e.g., \citeNP{BauHus09,MolDol10,NeaAgg06}).  In all, the tests have
advantageous properties that should be useful in practice.

>>>>

\title{Testing for measurement invariance with respect to an ordinal variable}
\threeauthors{Edgar C.\ Merkle}{Jinyan Fan}{Achim Zeileis}
\threeaffiliations{University of Missouri}{Auburn University}{Universit\"{a}t Innsbruck}

@@

As demonstrated in the simulations below, the ordinal test
statistics proposed above are sensitive to the measurement invariance
violations that an analyst would typically expect from an ordinal $V$.
In particular, due to computing cumulative sums in ${\bm B}(\hat {\bm \theta})$,
violations that occur as we move along the levels of $V$ can be captured
well. This includes abrupt shifts in the parameters $\bm \theta$ at a
certain level of $V$ as well as smooth increases/decreases in the parameters.
Taking a maximum over the $k$~parameters as in $\mathit{WDM}_o$ will be
more sensitive to changes that occur only in one out of many
parameters, 
while $\max \mathit{LM}_o$ will be more sensitive to changes occurring
in several (or even all of the) parameters simultaneously.
Moreover, the test statistics are rather insensitive to anomalies in a small
number of categories of $V$ that are
unrelated to the ordering of $V$.  This is especially relevant to situations
where the analyst has a large sample size, where the usual
likelihood ratio test is notoriously sensitive to minor parameter
instabilities (e.g., \citeNP{Ben80}).

@@

In this simulation, we demonstrate that the proposed test statistics are
sensitive to ordinal measurement invariance violations, moreso than
traditional statistics.  We generate data from a two-factor,
six-indicator model, with a measurement invariance violation occurring
in the unique variance parameters.  We
use the proposed test statistics to test for measurement invariance
simultaneously across the unique variances, which is similar to
comparing a tau-equivalent model to a congeneric model (e.g.,
\citeNP{VanLan00}).

We also compute two statistics that treat the ordinal
auxiliary variable as categorical.  These include the likelihood ratio
test of measurement 
invariance in the six unique variance parameters, along with the
unordered LM test from~\eqref{eq:lmuo}.

@@

\section{Simulation 2: Minor Anomalies and Large $n$}

@@

\subsection{Results}
Simulation results are presented in Figure~\ref{fig:sim2res}.  It is
observed that results are very consistent across the sample sizes
tested, implying that ``practical infinity'' is reached for this model
by $n=1200$.   We also observe a negative 
relationship between power and $m$; this is because the measurement
invariance violation occurred at only one level of the auxiliary
variable.  As $m$ increases (and $n$ is held
constant), the number of individuals violating measurement invariance
therefore decreases.  As a result, power to detect the violation decreases with
increasing $m$.

@@

Taken together, the results from Simulation 1 and Simulation
2 provide evidence that the $\text{LM}_o$ statistic should be
preferred to the $\text{WDM}_o$ statistic for simultaneously assessing
measurement invariance across multiple model parameters.  
\readme{This is not surprising given that the models
have quite a few parameters and we change many of them. We could 
try to set up a simulation where only a single out of many parameters
changes where the WDM should perform better. Also,
in models that are simpler than most CFA models, the WDM may be more useful.
I'm not sure how much we want to comment on this or illustrate this.}
\readme{Achim: I decided not to set up a case where a single parameter
changes, because measurement invariance researchers usually test
multiple parameters simultaneously.  For now, I added the phrase "across
multiple model parameters" above as a brief reference to this issue.}
The $\text{LM}_o$ statistic has higher power to detect
ordinal violations, and its power to detect non-ordinal violations was
similar to that of $\text{WDM}_o$ when the violation magnitude was
small (e.g., for $d \leq 1.5$).  The $\text{LM}_o$ statistic is
advantageous because it can make use of invariance violations that
simultaneously 
occur in multiple parameters, whereas the $\text{WDM}_o$ focuses only
on the parameter with the largest invariance violation.  
Thus, the statistics are likely to exhibit
similar performance if only a single model parameter violated
measurement invariance.

@@

\begin{figure}
\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
  the $\text{LM}_o$ statistic (right panel), arising from the GQ-6 data.}
\label{fig:gq6res}
\setkeys{Gin}{width=\textwidth}
<<gq6res, fig=TRUE, height=3, width=6>>=
par(mfrow=c(1,2), cex=.65)

@@

Plots representing the statistics' fluctuations across levels of age
group are displayed in Figure~\ref{fig:gq6res}.  The left panel
displays the process associated with $\text{WDM}_o$, while the right
panel displays the process associated with $\text{LM}_o$.  The final
age group (17-19 years) is not displayed, because the statistics
associated with this final age group are always zero (scores are
summed across all individuals in the sample at this point).
Critical values are displayed as horizontal red lines, so that the
hypothesis of invariance is rejected if the process ever crosses the
red line.
It is observed that both statistics generally increase with age, with
$\text{WDM}_o$ being largest in the 15-year group and $\text{LM}_o$
being largest in the 16-year group.  The $\text{WDM}_o$ panel displays
(a function of) the largest
cumulative sum associated with a single parameter, while the
$\text{LM}_o$ panel aggregates across the tested parameters.  The
differing pattern of values for the 15- and 16-year-olds can be taken
as evidence that a single factor loading is relatively unstable for the
15-year-olds, but this is not a general trend across all the tested
loadings.

@@

\begin{figure}
\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
  the $\text{LM}_o$ statistic (right panel), arising from the GAC data.}
\label{fig:gacres}
\setkeys{Gin}{width=\textwidth}
<<gacres, fig=TRUE, height=3, width=6>>=
par(mfrow=c(1,2), cex=.65)

plot(gefp2, functional=ordwmax(gefp2), axes=FALSE, main="",
     xlab="Age group", ylab=expression("WDM"[o]))
axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
axis(2)

plot(gefp2, functional=ol2bb2.fun, axes=FALSE, ylim=c(0,115), main="",
     xlab="Age group", ylab=expression("LM"[o]))
axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
axis(2)
@
\end{figure}

@@

Plots representing the statistics' fluctuations across levels of age
group are displayed in Figure~\ref{fig:gacres}.  The left panel
displays the process associated with $\text{WDM}_o$, while the right
panel displays the process associated with $\text{LM}_o$.  It is
observed that both processes are fully above the critical value,
implying the measurement invariance violation.  Additionally, both
processes peak at the 12-13 age group.  This implies that fluctuation 
in the variance parameters is largest within this age group, with the 
parameters differing between individuals younger and
older than 12 years.

\subsection{Summary}
The application considered above shows that the ordinal test
statistics can provide useful information in situations where one
might question significant likelihood ratio test statistics.  While
researchers use rules of thumb to obtain decisions from other
alternative fit measures, the proposed statistics are proper tests of
the hypothesis of interest.  They can be used to either supplement or
replace the likelihood ratio test, depending upon the types of
measurement
invariance violations in which the researcher has a priori interest.
We further describe the issue of supplementing vs.\ replacing the
likelihood ratio test in the general discussion.

@@

Along with using likelihood ratio tests to study measurement
invariance, researchers may wish to treat ordinal $V$ as continuous
(especially if $V$ has many levels).  This treatment would allow the
researcher to employ moderated factor models
\cite{BauHus09,MolDol10,Pur02} or factor mixture models
\cite{DolMaa98,LubMut05}.  Under the moderated factor model approach,
$V$ is inserted directly into the factor analysis model and allowed to
have a linear relationship with model parameters.  Under the factor
mixture model approach, individuals are typically assumed to arise
from a small number of distinct factor analysis models.  The ordinal
variable $V$ could then be used to predict the probability that an
individual arises from each model.  These treatments of ordinal $V$ as
continuous are likely to be advantageous, especially if the levels of
$V$ are approximately equally-spaced and the relationship between $V$
and the measurement invariance violation is linear.  The approaches do
require models of greater complexity and may not be suitable for all
ordinal $V$, however, while that the methods we propose here are
generally suitable for ordinal $V$.

@@

\subsection{Summary}
The test statistics proposed in this paper have relatively-high power for
detecting measurement 
invariance violations that are monotonic with the ordinal variable,
and they have relatively-low power for detecting minor violations that are not
monotonic.  The former feature implies that the statistics are good at
detecting measurement invariance violations that are interpretable to
the researcher, while the latter feature implies that the statistics
are feasible in situations where the likelihood ratio test commonly
rejects $H_0$ in practice (e.g., \citeNP{Ben80}).  Furthermore, the
focal psychometric model does not have to be modified in any way,
which differs from approaches that may treat the ordinal variable as
continuous.  In all, the tests have
advantageous properties that should be useful in practice.
